---
title: "DEGs Visualization"
author: "clagrabel"
date: "2024-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading the necessary packages

```{r}
#BiocManager::install("apeglm", force=TRUE)
library(apeglm)

pacman::p_load(tidyverse, DESeq2, ggVennDiagram, UpSetR, plotly, ggrepel, scico)
```

# Loading the data

```{r}
# DEG object
dds<-readRDS("./results/deseq2_dds.rds")
dds_test<-readRDS("./results/deseq2_dds_test.rds")

# The list of DEG results
res<-readRDS("./results/deseq2_results_populations.rds")
test<-readRDS("./results/deseq2_results_test.rds")
summary(res)
summary(test)

```

# Venn Diagrams

Extract DEGs lists in order to visualize overlapping regulated genes in the comparisons of interest.


```{r}
# List of DEGs

# Creating a function to extract the lists of genes repeteadly: all DEGs, up-regulated DEGs and down-regulated DEGs.

extract_degs<-function(x) {
  return(
    x %>%
      as_tibble(rownames = "gene") %>%
      filter(padj<0.05) %>%
      pull(gene)
  )
}

extract_up<-function(x) {
  return(
    x %>%
      as_tibble(rownames = "gene") %>%
      filter(padj<0.05) %>%
      filter(log2FoldChange>0) %>%
      pull(gene)
  )
}

extract_down<-function(x) {
  return(
    x %>%
      as_tibble(rownames = "gene") %>%
      filter(padj<0.05) %>%
      filter(log2FoldChange<0) %>%
      pull(gene)
  )
}

# Extracting all differentially expressed genes from all the comparisons stored in the list of DESeqResults.

all_degs<-lapply(res, FUN=extract_degs)
str(all_degs)

up_degs<-lapply(res, FUN=extract_up)
down_degs<-lapply(res, FUN=extract_down)

```


```{r eval=FALSE, include=FALSE}
#My way of doing venn diagrams

library(ggvenn)

venn_data <- list(nombre1=vector1,
                  nombre2=vector2,
                  nombre3=vector3)

venn_palette <- c("#e35023","#148733","#058dfc")

ggvenn(venn_data, columns = c("nombre1","nombre2","nombre3"),stroke_size = 0.3,
       fill_color = venn_palette, stroke_color="white", show_percentage=F,
       fill_alpha=0.6, set_name_color = venn_palette, text_size=6)
```


# Venn Diagrams

## Regulated genes in High and Low water in both Regions

It is interesting to check both up and down-regulated genes in each region in the different treatment conditions (high and low water).
This Venn Diagram would bring light over genes that are regulated in the Southern region which are not in the Central region and viceversa.


```{r}
all_degs[c("lS_hS","lC_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") + 
  ggtitle("DEGs in Southern Spain vs. DEGs in Central Spain")
```


## Regulated genes for different populations belonging to the same region

When looking at these diagrams, we expect to obtain the genes that are overlapping or not in the different populations belonging to the same region.

### Central Region

```{r}
all_degs[c("hBui_lBui","hCan_lCan", "hTur_lTur")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") + 
  ggtitle("DEGs in the populations from Central Spain")
```


### Southern Region

```{r}
all_degs[c("hEsp_lEsp","hLla_lLla", "hJab_lJab")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") + 
  ggtitle("DEGs in the populations from Southern Spain")
```


## Up and Down-regulated genes


```{r}
up_degs[c("lS_hS","lC_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") +
  ggtitle("Up-regulated genes South and Central")
```

```{r}
down_degs[c("lS_hS","lC_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") +
  ggtitle("Down-regulated genes South and Central")
```



# Test for all populations except Llano

## Regulated genes in High and Low water in both Regions

It is interesting to check both up and down-regulated genes in each region in the different treatment conditions (high and low water).
This Venn Diagram would bring light over genes that are regulated in the Southern region which are not in the Central region and viceversa.


```{r}
all_degs_test<-lapply(test, FUN=extract_degs)
str(all_degs_test)

up_degs_test<-lapply(test, FUN=extract_up)
down_degs_test<-lapply(test, FUN=extract_down)
```


```{r}
all_degs_test[c("test_lS_hS","test_lC_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") + 
  ggtitle("DEGs regarding water level in Southern Spain vs. DEGs in Central Spain")
```

Regarding water level response, there are 53 DEGs in Central Spain and 27 in Southern Spain. There are no common differentially expressed genes between Southern and Central regions.


## Up and Down-regulated genes


```{r}
up_degs_test[c("test_lS_hS","test_lC_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") +
  ggtitle("Up-regulated genes regarding water level South and Central")
```

Out of the DEGs, there are 48 genes that are up-regulated in Central Spain versus 11 in the South.


```{r}
down_degs_test[c("test_lS_hS","test_lC_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") +
  ggtitle("Down-regulated genes regarding water level South and Central")
```

However, the amount of down-regulated genes in Southern Spain is higher than in Central Spain.

### Regulated genes between regions for the same water level

```{r}
all_degs_test[c("test_lS_lC","test_hS_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow") + 
  ggtitle("DEGs in Southern Spain vs. DEGs in Central Spain")
```

Still, the larger effect is due to the region and not the water level. There is a higher amount of DEGs in the South compared to the Center, and there is a large number of common shared DEGs between regions.


## Venn Diagram all comparisons

From this diagram we can extract information about common DEGs for different comparisons.

```{r}
str(all_degs_test)

all_degs_test[c("test_lS_hS","test_lC_hC", "test_lS_lC", "test_hS_hC")] %>%
  ggVennDiagram(edge_size = 0) +
  scale_fill_scico(palette = "batlow")
```


```{r}
list_test<-all_degs_test[c("test_lS_hS","test_lC_hC", "test_lS_lC", "test_hS_hC")]

upset(fromList(list_test),
      number.angles = 0, point.size = 3, line.size = 1,
      sets.x.label = "Number of DEGs",
      set_size.show = TRUE,
      set_size.scale_max = max(sapply(list_test, length)), 
      text.scale = c(1.2, 1.2, 1.2, 1.2, 1.5, 1.5),
      order.by=c("degree","freq"))
```